---
title: "Model Evaluation"
subtitle: "Considerations for Time-to-Event Studies"
author: "Daniel Saggau"
date: "11/15/2020"
output:
  ioslides_presentation: default
  slidy_presentation: default
  powerpoint_presentation: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Overview

* Introduction to Time to Event Data
* Introduction to classical Model Evaluation Tools 
* Integrated Brier Score
* Concordance-Index
* Discussion 
* R- Implementation
* Advancements & Novel Considerations
* Recommendation

## Time-to Event Studies 

- Analysis working with right censored data
- Highly relevant for clinicians in the field of medical statistics e.g. looking at when a patient dies or when he gets a disease (clinical/epidemiological studies)
- In Economics/finance e.g. to examine when a subject/borrower will default or when a subject will find/lose a job 
- Operations research to predict the time a machine will break 

## Basic structure 

* Time T and Survival S
* Hazard h(t,x) to risk r(x)
* Capital H is the cumulative hazard
* non-parametric hazard models (KM) vs.semi-parametric proportional hazard model
* From hazard to cumulative hazard to survival 
* Survival Probability

## Model Evaluation - Considerations

(1) *What type of study are we dealing with?*

**Diagnostic vs. Prognostic Study**

(2) *What are the components of our model evaluation metric?*

**Discrimination**:
How well are we able to classify subjects ? Are we able to correctly discriminate between e.g. sick and healthy patients ?

**Calibration**:
How concise is our prediction accuracy ?
More recently also:

**Clinical Usefulness**: 
Does it make sense to implement our model in the first place? Will our model create more benefits than harm? 

## Classical Model Evaluation Tools 

**Classification Tasks**

* Brier Score (probability from true class label)
* Brier Score for ()
* Mis-classification Error rate (rate of incorrect classification)
* ROC (receiver operating characteristics)
* ACC (rate of correct classifications)

**Regression Tasks**

* Mean Absolute Error (mean absolute error)
* Mean Squared Error
* Brier Score 
* R^2 

### Brier Score 

* Specific Point in time 

Brier Score 1 

Brier Score 2 


### Confusion Matrix

**Sensitivity**:

* deals with values above the threshold among the subject group which do endure an event 
* Another common name for Sensitivity is the true positive rate.

$TPF = \frac{TP}{TP+FN}$

**Specificity**:

* deals with false negatives, hence patients with a disease we classify as not having any diseases
* Another name for specificity is the true negative rate 

$TNR = \frac{TN}{TN+FP}$

## Brier Score - Methods

## ROC/AUC/Concordance Statistics - Methods 

## Why cant we use traditional model evaluation tools for time to event studies?

* Working with censored data
``
* Right censored data (event after follow up) vs. left censored data (event was not recorded when it occured intially)
* We need to estimate survival of patients without having data on e.g. death
* Also, we need to provide measure over time

Early approaches:
 - excluding subjects with right censored data and only evaluate on the complete data

## From ROC to C-Statistic to C-index

- Advancement of ROC/AUC
- Further modification leads to c-index
- concordance pairs divided
- Harell's C

## c-index 

* studying pairs of subjects
* 
* addressing right censored data via inverse of the probability of censoring weighted estimate (of concordance probability)
* kendall's tau 
* Summary measure (over all time) based on the AUC  

$C-index = \frac{\Delta_j \times \sum_{i,j}1_{Ti>Tj} \times 1_{\eta_i>\eta_j}} {\Delta_j \times \sum_{i,j}1_{Ti>Tj}}$

* Where 1 are indicator-functions: 

## mlr3 Proba Applications: 

* van Houwelingen’s Alpha Calibration
* van Houwelingen’s Beta Calibration
* Integrated Graf Score
* Integrated Log Loss
* Log Loss

Further measures via survAUC package:

* Uno's AUC/TPR/TNR
* Song and Zhou’s AUC/TNR/TPR
* Chambless and Diao’s AUC
* Hung and Chiang’s AUC


```{r}
library(prodlim)
library(survival)
library(randomForestSRC)
 set.seed(13)
 dat <- SimSurv(500)
 # fit three different Cox models and a random survival forest
 cox12 <- coxph(Surv(time,status)~X1+X2,data=dat,x=TRUE,y=TRUE)
 cox1 <- coxph(Surv(time,status)~X1,data=dat,x=TRUE,y=TRUE)
 cox2 <- coxph(Surv(time,status)~X2,data=dat,x=TRUE,y=TRUE)
 rsf1 <- rfsrc(Surv(time,status)~X1+X2,data=dat,ntree=15,forest=TRUE)

A1  <- pec::cindex(list("Cox X1"=cox1, "RSF"=rsf1),
formula=Surv(time,status)~X1+X2,
data=dat,
eval.times=10)
ApparrentCindex  <- pec::cindex(list("Cox X1"=cox1,
"Cox X2"=cox2,
"Cox X1+X2"=cox12,
"RSF"=rsf1),
formula=Surv(time,status)~X1+X2,
data=dat,
eval.times=seq(1,15,1))
  print(ApparrentCindex)
  plot(ApparrentCindex)
```

## IBS

* called cumulative predictive error curves == continuous ranked probability score (crps)
* area under the prediction error curve
* Integral over all points in time to get one summary value henceforth called "integrated" BS
* able to build a R^2 like measure where we divide MSE of a model with a different MSE of reference model 

For the Individual: 

$$L(S,t|t^*) = [(S(t^*)^2)I(t)\leq t^*, \delta =1)(\frac{1}{G(t)})]$$ \newline
$$+ [((1-S(t^*))^2)I(t>t^*)(\frac{1}{G(t^*)})]$$


* Where L is a loss function of the S(the probability that the event of interest has not taken place yet) and time
* t is the time of the event (death) and t* the time before death
* G(t) is the P(C>t), so where the censored time is longer than the time (in mlr3proba via survfit == KM Estimate)

* When selecting integrated == FALSE then we looking at specific time 

## For the population mean:

$$
L(S,t|t^*) = \frac{1}{N}\displaystyle\sum^N_{i=1}L(S_i,t_i|t^*) \tag{9}
$$

### Mean Population: 

$$L(S,t|t^*) = \frac{1}{NT}\displaystyle\sum^N_{i=1}\displaystyle\sum ^T_{j=1}L(S_i,t_i|t^*)$$

* N = Number of observations 
* S_i is the predicted survival function 

## If we take

```{r, include =F}
set.seed(130971)
library(prodlim)
library(survival)
library(pec)
```
 
```{r}
set.seed(18713)
dat=SimSurv(100)
pmodel=coxph(Surv(time,status)~X1+X2,data=dat,x=TRUE,y=TRUE)
perror=pec(list(Cox=pmodel),Hist(time,status)~1,data=dat)
## cumulative prediction error
# crps(perror,times=1) # between min time and 1
## same thing:
y<- ibs(perror,times=1) # between min time and 1 crps(perror,times=1,start=0) # between 0 and 1 crps(perror,times=seq(0,1,.2),start=0) # between 0 and seq(0,1,.2)
plot(perror)
calPlot(pmodel)
```

```{r}
dat <- SimSurv(1000)

Models <- list("Cox.X1"=coxph(Surv(time,status)~X1,data=dat,x=TRUE,y=TRUE),
               "Cox.X2"=coxph(Surv(time,status)~X2,data=dat,x=TRUE,y=TRUE),
               "Cox.X1.X2"=coxph(Surv(time,status)~X1+X2,data=dat,x=TRUE,y=TRUE))

PredError <- pec(object=Models,
                 formula=Surv(time,status)~X1+X2, data=dat,
                 exact=TRUE, cens.model="marginal", splitMethod="none",
                 B=0,
                 verbose=TRUE)

print(PredError,times=seq(1,15,1))
summary(PredError)
```

## Plotting the whole damn thing

```{r}
plot(PredError,xlim=c(0,20), ylim = c(0,1))
```

## Discussion 

* Integrated Brier Score accounts for both calibration and discrimination
* Irrespective, neither model accounts and leaves room for improvement 

## Literature

Introduction: 

- Steyerberg, E. W., Vickers, A. J., Cook, N. R., Gerds, T., Gonen, M., Obuchowski, N., ... & Kattan, M. W. (2010). Assessing the performance of prediction models: a framework for some traditional and novel measures. Epidemiology (Cambridge, Mass.), 21(1), 128.

Brier Score: 

- Graf, E., Schmoor, C., Sauerbrei, W., & Schumacher, M. (1999). Assessment and comparison of prognostic classification schemes for survival data. Statistics in medicine, 18(17‐18), 2529-2545.

## Advanced Literature on C-index:

- Antolini, L., Boracchi, P., & Biganzoli, E. (2005). A time‐dependent discrimination index for survival data. Statistics in medicine, 24(24), 3927-3944.

- Uno, H., Cai, T., Pencina, M. J., D'Agostino, R. B., & Wei, L. J. (2011). On the C‐statistics for evaluating overall adequacy of risk prediction procedures with censored survival data. Statistics in medicine, 30(10), 1105-1117.

## Comparative Study: 

- Kattan, M. W., & Gerds, T. A. (2018). The index of prediction accuracy: an intuitive measure useful for evaluating risk prediction models. Diagnostic and prognostic research, 2(1), 7.

