---
title: "Model Evaluation"
subtitle: "Considerations for Time-to-Event Studies"
author: "Daniel Saggau"
date: "11/12/2020"
output:
  beamer_presentation:
    theme: "Madrid"
    colortheme: "spruce"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T)
library("mlr3learners")
library("mlr3")
library("mlr3proba")
library("survival")
library("ranger")
library("mlr3viz")
library("survAUC")
library("prodlim")
library("pec")
```

## Overview

1) Time to Event Studies

* What differentiates a TTE Study from other studies?

2) Classical Model Evaluation: Brier Score and AUC

* What are classical model evaluation tools & why can't we use them? 

3) TTS Model Evaluation: IBS and c-index

* How do these methods address the limitations of classical methods? 

4) Discussion

* What measure is most useful for machine learning in TTS? 

5) Further Considerations

* What other methods are coming? 

## Time-to Event Studies 

- Analysis working with (right) censored data
- Right censored data (event after follow up) vs. left censored data (event was not recorded when it occurred initially)
- Highly relevant for clinicians in the field of medical statistics e.g. looking at when a patient dies or when he gets a disease (clinical/epidemiological studies)
- In Economics/Finance e.g. to examine when a subject/borrower will default or when a subject will find/lose a job 
- Operations research to predict the time a machine will break 

## Basic Notations & Concepts

* Time T and Survival S
* From hazard to cumulative hazard to survival 
* Hazard h(t,x) is the eminent probability of death a specific point in time 
* Capital H is the cumulative hazard
* non-parametric hazard models (KM) vs.semi-parametric proportional hazard model

## Classical Model Evaluation Tools for Classification Tasks

(1) **Diagnostic vs. Prognostic Study**
(2) **What elements do we consider?**

* Discrimination: Are we able to correctly discriminate between e.g. sick and healthy patients ?
* Calibration: How concise is our prediction accuracy ?
* Clinical Usefulness: Will our model create more benefits than harm? 

(3) Working with *Label* vs. working with *Probability*

* Brier Score (probability from true class label)
* AUC (label based error measure via specificity and sensitivity)

## Brier Score 

Based on loss function.
Other loss measures are the log loss or the integrated log loss.

MSE for Regression (L2 Loss):

$MSE= \frac{1}{n}\sum^n_{i=1}(y^{(i)})-\hat{y}^{(i)})^2$

Where: the $MSE \in [0;\infty)$

The Brier Score is the MSE for Classification: 

$BS= \frac{1}{n}\sum^n_{i=1}(\hat{\pi}(x^{(i)})-y^{(i)})^2$

The general version of the brier score looks at a specific point in time
We can plot this brier score via prediction error curves (pec)

## Confusion Matrix

**Sensitivity** or: true positive rate

* deals with values above the threshold among the subject group which do endure an event 

$TPF = \frac{TP}{TP+FN}$

**Specificity** or: true negative rate 

* deals with false negatives, hence patients with a disease we classify as not having any diseases

$TNR = \frac{TN}{TN+FP}$

## Why cant we use traditional model evaluation tools for time to event studies?

* Working with censored data
* Account for time dependent covariates 

Early approaches:
 - excluding subjects with right censored data and only evaluate on the complete data

## From AUC to Harell's C-index to time dependent C-index

* Advancement from AUC 
* Rank correlation measure but still have to deal with censoring
* studying concordance (~consistency) and discordance (~inconsistency) pairs
* Core difference is that

Intuitively speaking the difference between AUC and c-index is as follows: 

$AUC = \mathrm{Prob(Risk_t(i)> Risk_t(j)| i\: has\: event\: before\: t\: and \: j \: has\: event\: after \: t})$
$C = \mathrm{Prob(Risk_t(i)> Risk_t(j)| i\: has\: event\: before\: t)}$

Further, we could relabel those terms for the C as: 

$\frac{\#Concordant Pairs}{\#Concordant Pairs + \#Discordant Pairs}$

Mathematically, we can define e.g. the C for time dependent covariates as: 

$C ^{td} = \frac{Prob( z(X_i) > z(X_j) \& T_i < T_j \& D_i = 1}{Prob(Ti<Tj | D_i = 1)}$

## c-index 

**How to deal with censoring:**
* addressing right censored data via inverse of the probability of censoring weighted estimate (of concordance probability)
* Kendall rank correlation coefficient test as inspiration
* Summary measure (over all time) based on the AUC  

$C-index = \frac{\Delta_j \times \sum_{i,j}1_{Ti>Tj} \times 1_{\eta_i>\eta_j}} {\Delta_j \times \sum_{i,j}1_{Ti>Tj}}$


## IBS

* called cumulative predictive error curves == continuous ranked probability score (crps)
* area under the prediction error curve
* Integral over all points in time to get one summary value henceforth called "integrated" BS
* able to build a R^2 like measure where we divide MSE of a model with a different MSE of reference model 
* Where L is a loss function of the S(the probability that the event of interest has not taken place yet) and time
* t is the time of the event (death) and t* the time before death
* G(t) is the P(C>t), so where the censored time is longer than the time (in mlr3proba via survfit == KM Estimate)
* When selecting integrated == FALSE then we looking at specific time 

## Mean population

**Mean Population:without Integration**

$$
L(S,t|t^*) = \frac{1}{N}\displaystyle\sum^N_{i=1}L(S_i,t_i|t^*) \tag{9}
$$

**Mean Population: with Integration** 

$$L(S,t|t^*) = \frac{1}{NT}\displaystyle\sum^N_{i=1}\displaystyle\sum ^T_{j=1}L(S_i,t_i|t^*)$$

* N = Number of observations 
* S_i is the predicted survival function 

## Coding Settup
\footnotesize
```{r}
set.seed(123)
dat=SimSurv(10000)
models <-  list("Cox.X1"=coxph(Surv(time,status)~X1,
                      data=dat, x=TRUE,y=TRUE),
               "Cox.X2"=coxph(Surv(time,status)~X2,
                      data=dat,x=TRUE,y=TRUE),
               "Cox.X1.X2"=coxph(Surv(time,status)~X1+X2,
                      data=dat,x=TRUE,y=TRUE))
```

## Defining the prediction error based on the brier score
\footnotesize
```{r}
perror <- pec(object=models,
                 formula=Surv(time,status)~1,  
              data=dat, 
                 exact=TRUE, cens.model="marginal",
              splitMethod="none", 
                 B=0,# number boostrap samples 
                 verbose=TRUE)
```

## Calibration Plot

```{r}
calPlot(models)
```

## Summary Prediction Error Curve
\tiny
```{r}
summary(perror,times= quantile(dat$time[dat$status==1], c(.25, .5, .75,1)))
```

## Plotting prediction error

```{r}
plot(perror,xlim=c(0,25), ylim = c(0,0.8))
```

## Cumulative Prediction Error 
\footnotesize
```{r, out.width=1}
#crps(object, models, what, times, start)
crps(perror,times= quantile(dat$time[dat$status==1], c(.25, .5, .75, 1)))
# ibs(perror,times= quantile(dat$time[dat$status==1], c(.25, .5, .75, 1)))
```

## cindex Implementation 

### Components of the c-index function
cindex(*object, formula, cens.model,data, eval.times, cause, data, splitMethod, B,M...*)

* **formula** is our survival formula (Surv(time,status)~x1+x2 for cens.model="cox" or Surv(time,status)~1 for cens.model ="marginal")
* **cens.model** is our method for estimating the inverse probability of censoring weights (e.g. cox, marginal, nonpar)
* **splitMethod** is the internal validation design,  B the number of boostrap samples & M the size of the boostrap sample
* **cause** used for competing risks (default is the first state of the response)

\tiny
```{r}
cindex = cindex(models, formula = Surv(time,status) ~ 1,
     cens.model="marginal", data = dat,
      eval.times= quantile(dat$time[dat$status==1], c(.25, .5, .75,1)))
```

## c-index plot

```{r, fig.width= 6, fig.height=4}
plot(cindex,xlim=c(0,24), ylim =c(0.55,1))
```


## mlr3Proba 

Methods based on the loss function:

* Integrated Graf Score (other Name for IBS based on Author Graf)
* Integrated Log Loss (surpress scale of variation)
* Log Loss  (censored data ignored)

Further measures via survAUC package:

* Uno's AUC 
* Song and Zhou’s AUC


## mlr3Proba Example

```{r, include =F}
TaskSurv$new(id = "interval_censored", backend = survival::bladder2[,-c(1, 7)],
             time = "start", time2 = "stop", type = "interval2")
task = tsk("rats")
learners = lrns(c("surv.coxph", "surv.kaplan", "surv.ranger"))
measure = msr("surv.graf") # for c-index you can use surv.cindex
bmr = benchmark(benchmark_grid(task, learners, rsmp("cv", folds = 3)))
bmr$aggregate(measure)
```

```{r, fig.width= 3.5, fig.height=3}
autoplot(bmr, measure = measure)
```

## Discussion 

* c-index has gained popularity because of it's interpretability
* Integrated Brier Score accounts for both calibration and discrimination
* Irrespective, neither model accounts and leaves room for improvement 
* IBS allows for differentiation of 'useless' and 'harmful'
* Estimators can be influenced by data
* Clinical consequences problematic 

## Novel Research

* Decision Curve Analysis (clinical consequences): plotting different exchange rates with the net benefit equation 
* Net Reclassification Improvement (clinical consequences)
* Other estimators like SVM estimators for the evaluations tools for the censored data
* IPA 
* Competing Risks

## Conclusion 

* There are various different modifications for model evaluation, neither being unconditionally superior 
* The Brier Score and the AUC are pivotal for many of these methods
* While there has been a lot of research on this topic, the debate is on going

## Literature and Recommendations 

Introduction:

* Steyerberg, E. W., Vickers, A. J., Cook, N. R., Gerds, T., Gonen, M., Obuchowski, N., ... & Kattan, M. W. (2010). Assessing the performance of prediction models: a framework for some traditional and novel measures. Epidemiology (Cambridge, Mass.), 21(1), 128.

Comparative Study:

* Kattan, M. W., & Gerds, T. A. (2018). The index of prediction accuracy: an intuitive measure useful for evaluating risk prediction models. Diagnostic and prognostic research, 2(1), 7.

## Use Cases:

https://rpubs.com/kaz_yos/survival-auc
https://datascienceplus.com/time-dependent-roc-for-survival-prediction-models-in-r/
https://rdrr.io/cran/pec/
https://adibender.github.io/pammtools/
